\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-25 paper/manuscript.qmd
abstract

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  As AI-generated content becomes increasingly indistinguishable from real stimuli, understanding how individuals form beliefs about the reality of what they perceive is both theoretically and practically urgent. This study introduces the Choice-Confidence (CHOCO) model as a tool to dissect the cognitive mechanisms underlying decision-making judgments. Conceptualized as a mixture of two Beta distributions, CHOCO simultaneously estimates the probability of selecting one category (e.g., "real" vs. "AI-generated") and the confidence associated with each choice. We apply this model to two datasets (N=141 and N=189) in which participants judged whether faces were real photographs or AI-generated images, finding that facial attractiveness systematically increased the likelihood of being judged as "real", particularly for male participants. We discuss how these findings relate to the broader effect of facial attractiveness, and suggest future directions for both the cognitive and psychometric application of the CHOCO framework.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness, simulation monitoring"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

**Study 1** aims to evaluate the CHOCO model’s ability to better capture subjective scale responses that (potentially) reflect an underlying discrete choice, in comparison to existing models such as the ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO provides improved model fit, 2) yields deeper insights into population-level effects than traditional approaches (sex differences in reality beliefs), and 3) allows for the reliable estimation of interpretable individual-level parameters through random effects. **Study 2** will apply this model to more subtle effects, such as the effect of perceived facial attractiveness on reality judgments, and test the ability to fit alternative data structures (scales with ordinal response options and mid-points). To this end, we analyze data from two separate studies in which participants judged whether a face image was AI-generated ("fake") or a real photograph.

@@

This study provides further evidence that attractive faces are more likely to be judged as real (*vs.* AI-generated), This effect was particularly strong and robust among male participants. For females, the effect was weaker in sample 1 but more pronounced in sample 2, which might partly reflect increased statistical power. An alternative possibility is that the second sample's experimental manipulation introduced conflicting cues (e.g., faces labelled "AI-generated") that could have disrupted intuitive heuristics, thereby increasing reliance on attractiveness as a diagnostic cue, particularly in ambiguous trials.
<!-- This could be tested by analyzing believed vs. non-believed items -->

>>>>

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  We introduce the Choice-Confidence (CHOCO) model as a tool to dissect the cognitive mechanisms underlying decision-making judgments, offering a better fit for data from subjective rating scales for which the two extremes correspond to discrete choices (e.g., "positive" *vs.* "negative", "true" *vs.* false", "real" *vs.* "AI-generated", ...), and for which the distance from the center might reflect a gradual process, such as confidence or certainty in that choice. Conceptualized as a mixture of two Beta distributions, CHOCO simultaneously estimates the probability of selecting one category (the right side of the scale over the left side) and the confidence associated with each choice. To demonstrate its usage, we apply this model to two datasets (N=141 and N=189) in which participants judged whether face images were real photographs or AI-generated images, finding that facial attractiveness increased the likelihood of being judged as "real", particularly for male participants. As AI-generated content becomes increasingly indistinguishable from real stimuli, understanding how individuals form beliefs about the reality of what they perceive is both theoretically and practically urgent. We discuss how these findings relate to the broader effect of facial attractiveness, and suggest future directions for both the cognitive and psychometric application of the CHOCO framework.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness, simulation monitoring, zero-and-one inflated Beta regression"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

**Study 1** aims to evaluate the CHOCO model’s ability to better capture subjective scale responses that (potentially) reflect an underlying discrete choice, in comparison to existing models such as the ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO provides improved model fit, 2) yields deeper insights into population-level effects than traditional approaches (sex differences in reality beliefs), and 3) allows for the reliable estimation of interpretable individual-level parameters through random effects. **Study 2** will apply this model to more subtle effects, such as the effect of perceived facial attractiveness on reality judgments, and test the ability to fit alternative data structures (scales with ordinal response options and mid-points). As a case-study illustration, we analyze data from two separate studies in which participants judged whether a face image was AI-generated ("fake") or a real photograph.

@@

This study provides further evidence that attractive faces are more likely to be judged as real (*vs.* AI-generated). This effect was particularly strong and robust among male participants. For females, the effect was weaker in sample 1 but more pronounced in sample 2, which might partly reflect increased statistical power. An alternative possibility is that the second sample's experimental manipulation introduced conflicting cues (e.g., faces labelled "AI-generated") that could have disrupted intuitive heuristics, thereby increasing reliance on attractiveness as a diagnostic cue, particularly in ambiguous trials.
<!-- This could be tested by analyzing believed vs. non-believed items -->

### 2025-06-25 paper/manuscript.tex
abstract

\abstract{As AI-generated content becomes increasingly indistinguishable
from real stimuli, understanding how individuals form beliefs about the
reality of what they perceive is both theoretically and practically
urgent. This study introduces the Choice-Confidence (CHOCO) model as a
tool to dissect the cognitive mechanisms underlying decision-making
judgments. Conceptualized as a mixture of two Beta distributions, CHOCO
simultaneously estimates the probability of selecting one category
(e.g., ``real'' vs.~``AI-generated'') and the confidence associated with
each choice. We apply this model to two datasets (N=141 and N=189) in
which participants judged whether faces were real photographs or
AI-generated images, finding that facial attractiveness systematically
increased the likelihood of being judged as ``real'', particularly for
male participants. We discuss how these findings relate to the broader
effect of facial attractiveness, and suggest future directions for both
the cognitive and psychometric application of the CHOCO framework. }

\keywords{subjective scales, choice confidence model, reality beliefs,
AI-generated faces, attractiveness, simulation monitoring}

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, opacityback=0, left=2mm, leftrule=.75mm, colback=white, breakable, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, arc=.35mm, rightrule=.15mm]

@@

\textbf{Study 1} aims to evaluate the CHOCO model's ability to better
capture subjective scale responses that (potentially) reflect an
underlying discrete choice, in comparison to existing models such as the
ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO
provides improved model fit, 2) yields deeper insights into
population-level effects than traditional approaches (sex differences in
reality beliefs), and 3) allows for the reliable estimation of
interpretable individual-level parameters through random effects.
\textbf{Study 2} will apply this model to more subtle effects, such as
the effect of perceived facial attractiveness on reality judgments, and
test the ability to fit alternative data structures (scales with ordinal
response options and mid-points). To this end, we analyze data from two
separate studies in which participants judged whether a face image was
AI-generated (``fake'') or a real photograph.

@@

This study provides further evidence that attractive faces are more
likely to be judged as real (\emph{vs.} AI-generated), This effect was
particularly strong and robust among male participants. For females, the
effect was weaker in sample 1 but more pronounced in sample 2, which
might partly reflect increased statistical power. An alternative
possibility is that the second sample's experimental manipulation
introduced conflicting cues (e.g., faces labelled ``AI-generated'') that
could have disrupted intuitive heuristics, thereby increasing reliance
on attractiveness as a diagnostic cue, particularly in ambiguous trials.

>>>>

\abstract{We introduce the Choice-Confidence (CHOCO) model as a tool to
dissect the cognitive mechanisms underlying decision-making judgments,
offering a better fit for data from subjective rating scales for which
the two extremes correspond to discrete choices (e.g., ``positive''
\emph{vs.} ``negative'', ``true'' \emph{vs.} false'', ``real''
\emph{vs.} ``AI-generated'', \ldots), and for which the distance from
the center might reflect a gradual process, such as confidence or
certainty in that choice. Conceptualized as a mixture of two Beta
distributions, CHOCO simultaneously estimates the probability of
selecting one category (the right side of the scale over the left side)
and the confidence associated with each choice. To demonstrate its
usage, we apply this model to two datasets (N=141 and N=189) in which
participants judged whether face images were real photographs or
AI-generated images, finding that facial attractiveness increased the
likelihood of being judged as ``real'', particularly for male
participants. As AI-generated content becomes increasingly
indistinguishable from real stimuli, understanding how individuals form
beliefs about the reality of what they perceive is both theoretically
and practically urgent. We discuss how these findings relate to the
broader effect of facial attractiveness, and suggest future directions
for both the cognitive and psychometric application of the CHOCO
framework. }

\keywords{subjective scales, choice confidence model, reality beliefs,
AI-generated faces, attractiveness, simulation monitoring, zero-and-one
inflated Beta regression}

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, leftrule=.75mm, left=2mm, rightrule=.15mm, colback=white, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, opacityback=0, arc=.35mm]

@@

\textbf{Study 1} aims to evaluate the CHOCO model's ability to better
capture subjective scale responses that (potentially) reflect an
underlying discrete choice, in comparison to existing models such as the
ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO
provides improved model fit, 2) yields deeper insights into
population-level effects than traditional approaches (sex differences in
reality beliefs), and 3) allows for the reliable estimation of
interpretable individual-level parameters through random effects.
\textbf{Study 2} will apply this model to more subtle effects, such as
the effect of perceived facial attractiveness on reality judgments, and
test the ability to fit alternative data structures (scales with ordinal
response options and mid-points). As a case-study illustration, we
analyze data from two separate studies in which participants judged
whether a face image was AI-generated (``fake'') or a real photograph.

@@

This study provides further evidence that attractive faces are more
likely to be judged as real (\emph{vs.} AI-generated). This effect was
particularly strong and robust among male participants. For females, the
effect was weaker in sample 1 but more pronounced in sample 2, which
might partly reflect increased statistical power. An alternative
possibility is that the second sample's experimental manipulation
introduced conflicting cues (e.g., faces labelled ``AI-generated'') that
could have disrupted intuitive heuristics, thereby increasing reliance
on attractiveness as a diagnostic cue, particularly in ambiguous trials.
